The Power of Two Random Choices\\ \large A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Träff, lecture ``Parallel Algorithms'', 2012 winter term at TU Wien.}

### 2014-02-09 article/article.tex
Use align environment for formulas

The fullest bin contains
\[
\Theta \left( \frac{\log n}{\log \log n} \right)
\] 
balls in the lightly loaded case w.h.p. Since the average load of the bins is 1, this is also a bound for the gap. 
%More accurately, the number of balls in the bin with the most balls is $\Gamma^{-1}\left(n\right)\left(1+O\left(\frac{1}{\log \Gamma^{-1}\left(n\right)}\right)\right)$ \cite{G91}. %\todo{really include Gonnet?}
In the \emph{heavily loaded case} the maximum load is 
\[
\frac{m}{n} + \Theta\left(\sqrt{\frac{m \log n}{n}}\right)
\] 
w.h.p.~\cite{RS98}.

@@

Thus, let $B$ be the set of bins $\{1,2...n\}$, $\Omega = B^{d}$ be the sample space and $\mathcal{F} = 2^{\Omega}$ be the set of measurable events. We are only interested in events that represent a set of bins of size $d$. Thus, let $\mathcal{F}_d = \{\mathcal{E} \in \mathcal{F}: \left\vert \mathcal{E} \right\vert = d\}$ and set $\Pr\left(\mathcal{E}  \right) = 0$  $\forall \mathcal{E} \in \mathcal{F} \setminus \mathcal{F}_d $. The three classes of algorithms are distinguished by the probability function $\Pr\left(\mathcal{E}\right)$ for events $\mathcal{E} \in \mathcal{F}_d$:
\begin{compactitem}
\item  \emph{Uniform and independent.} Choose each of the $d$ bins for a ball uniformly and independently at random from the set of bins $B$. This implies 
\[
\forall \mathcal{E} \in \mathcal{F}_d: \Pr\left(\mathcal{E}\right) = \frac{1}{n ^{d}}
\]
\item Class 2: \emph{Non-uniform and independent.} Choose the $i$th bin from $B$ independently at random according to some probability function
\[
\Pr: B \rightarrow \left[0,1\right]
\]

\item Class 3: \emph{Non-uniform and dependent.} Choose the $d$ bins for a ball from $\Omega$ at random:
\[
\Pr: B^{d} \rightarrow \left[0,1\right]
\]
\end{compactitem}

@@

\begin{theorem}
\label{theorem:greedy}
Consider any sequence of insertions and deletions s.t.~at most $hn$ balls are in the bins at any time. If the balls are placed into bins by the Greedy algorithm, at any time $t$ the maximum load is at most 
\[
\frac{m}{n} + \Theta\left(\frac{\log \log n}{\log d} \right)
\]
w.h.p.~at any time $t$ \cite{ABKU99} \cite{BCSV06}.

@@

\begin{theorem}
\label{theorem:agln}
If $n$ balls are placed into $n$ bins using the Always-go-left scheme, the number of balls in the fullest bin will be 
\[
\frac{\ln\ln n}{d \ln \Phi_d} \pm \Theta(1)
\]
w.h.p.
\end{theorem}

\begin{theorem}
\label{theorem:algm}
Consider any sequence of insertions and deletions s.t.~at most $hn$ balls are in the bins at any time. If the balls are placed into bins by the Greedy algorithm, at any time $t$ the maximum load is at most 
\[
\frac{\ln\ln n}{d \ln \Phi_d} + O(h)
\]
w.h.p.
\end{theorem}

@@

\[
F_d(k) = \sum_{i=1}^{d}F_d(k-i)
\]
Furthermore, define 
\[
\Phi_d = \lim_{k \rightarrow \infty} \sqrt[k]{F_d(k)}
\]
Notice that $F_2$ corresponds to the usual Fibonacci numbers, while $\Phi_2 $ corresponds to the golden ratio \cite{Knuth73}, which is generalized to $\Phi_d$, for which 
\[
i, j \in \mathbb{N}, i < j \Rightarrow \Phi_i < \Phi_j
\]
and 
\[
\lim_{d\rightarrow \infty} \Phi_d = 2
\]
hold.

@@

An edge\footnote{where $v$ is the $i$'th child of $u$} $(u, v)$ is activated with probability at most $d/n$, which is the probability that the $i$'th location of the ball $b_u = \mathrm{ball}(u)$ points to a particular location of the ball $b_v = \mathrm{ball}(v)$ times the number of locations $d$. This implies a probability of at most
\[
p_e = \left(\frac{d}{n}\right)^{k-1}
\] 
or the event that all $k-1$ edges of a witness tree are activated. Note that  the assumption that all balls are distinct (Assumption \ref{assumption:independence}) implies that all edge events are independent.

The probability that a leaf event is activated is bounded above by $3^{-d}$. Recall that by the definition of leaf events, each of the $d$ locations of the ball $b_{vl} = \mathrm{ball}(v_l)$ has to point to a bin with at least 3 additional balls in it. The number of those bins is bounded by $\lfloor n/3 \rfloor$ at all time steps $t$, implying a probability of at most 
\[
\frac{n/3}{n} = \frac{1}{3}
\end{align}
that one such bin is chosen. This implies a probability of at most $3^{-d}$ that all $d$ bins are chosen among those with at least 3 additional bins in them. Thus, the probability that all leaf events are activated is at most $p_l = 3^{-d\cdot l}$, where $l = d^{L}$ is the number of leaves in a witness tree.

>>>>

The fullest bin contains
\begin{align}
\Theta \left( \frac{\log n}{\log \log n} \right)
\end{align} 
balls in the lightly loaded case w.h.p. Since the average load of the bins is 1, this is also a bound for the gap. 
%More accurately, the number of balls in the bin with the most balls is $\Gamma^{-1}\left(n\right)\left(1+O\left(\frac{1}{\log \Gamma^{-1}\left(n\right)}\right)\right)$ \cite{G91}. %\todo{really include Gonnet?}
In the \emph{heavily loaded case} the maximum load is 
\begin{align}
\frac{m}{n} + \Theta\left(\sqrt{\frac{m \log n}{n}}\right)
\end{align} 
w.h.p.~\cite{RS98}.

@@

Thus, let $B$ be the set of bins $\{1,2...n\}$, $\Omega = B^{d}$ be the sample space and $\mathcal{F} = 2^{\Omega}$ be the set of measurable events. We are only interested in events that represent a set of bins of size $d$. Thus, let $\mathcal{F}_d = \{\mathcal{E} \in \mathcal{F}: \left\vert \mathcal{E} \right\vert = d\}$ and set $\Pr\left(\mathcal{E}  \right) = 0$  $\forall \mathcal{E} \in \mathcal{F} \setminus \mathcal{F}_d $. The three classes of algorithms are distinguished by the probability function $\Pr\left(\mathcal{E}\right)$ for events $\mathcal{E} \in \mathcal{F}_d$:
\begin{compactitem}
\item  \emph{Uniform and independent.} Choose each of the $d$ bins for a ball uniformly and independently at random from the set of bins $B$. This implies 
\begin{align}
\forall \mathcal{E} \in \mathcal{F}_d: \Pr\left(\mathcal{E}\right) = \frac{1}{n ^{d}}
\end{align}
\item Class 2: \emph{Non-uniform and independent.} Choose the $i$th bin from $B$ independently at random according to some probability function
\begin{align}
\Pr: B \rightarrow \left[0,1\right]
\end{align}

\item Class 3: \emph{Non-uniform and dependent.} Choose the $d$ bins for a ball from $\Omega$ at random:
\begin{align}
\Pr: B^{d} \rightarrow \left[0,1\right]
\end{align}
\end{compactitem}

@@

\begin{theorem}
\label{theorem:greedy}
Consider any sequence of insertions and deletions s.t.~at most $hn$ balls are in the bins at any time. If the balls are placed into bins by the Greedy algorithm, at any time $t$ the maximum load is at most 
\begin{align}
\frac{m}{n} + \Theta\left(\frac{\log \log n}{\log d} \right)
\end{align}
w.h.p.~at any time $t$ \cite{ABKU99} \cite{BCSV06}.

@@

\begin{theorem}
\label{theorem:agln}
If $n$ balls are placed into $n$ bins using the Always-go-left scheme, the number of balls in the fullest bin will be 
\begin{align}
\frac{\ln\ln n}{d \ln \Phi_d} \pm \Theta(1)
\end{align}
w.h.p.
\end{theorem}

\begin{theorem}
\label{theorem:algm}
Consider any sequence of insertions and deletions s.t.~at most $hn$ balls are in the bins at any time. If the balls are placed into bins by the Greedy algorithm, at any time $t$ the maximum load is at most 
\begin{align}
\frac{\ln\ln n}{d \ln \Phi_d} + O(h)
\end{align}
w.h.p.
\end{theorem}

@@

\begin{align}
F_d(k) = \sum_{i=1}^{d}F_d(k-i)
\end{align}
Furthermore, define 
\begin{align}
\Phi_d = \lim_{k \rightarrow \infty} \sqrt[k]{F_d(k)}
\end{align}
Notice that $F_2$ corresponds to the usual Fibonacci numbers, while $\Phi_2 $ corresponds to the golden ratio \cite{Knuth73}, which is generalized to $\Phi_d$, for which 
\begin{align}
i, j \in \mathbb{N}, i < j \Rightarrow \Phi_i < \Phi_j
\end{align}
and 
\begin{align}
\lim_{d\rightarrow \infty} \Phi_d = 2
\end{align}
hold.

@@

An edge\footnote{where $v$ is the $i$'th child of $u$} $(u, v)$ is activated with probability at most $d/n$, which is the probability that the $i$'th location of the ball $b_u = \mathrm{ball}(u)$ points to a particular location of the ball $b_v = \mathrm{ball}(v)$ times the number of locations $d$. This implies a probability of at most
\begin{align}
p_e = \left(\frac{d}{n}\right)^{k-1}
\end{align} 
or the event that all $k-1$ edges of a witness tree are activated. Note that  the assumption that all balls are distinct (Assumption \ref{assumption:independence}) implies that all edge events are independent.

The probability that a leaf event is activated is bounded above by $3^{-d}$. Recall that by the definition of leaf events, each of the $d$ locations of the ball $b_{vl} = \mathrm{ball}(v_l)$ has to point to a bin with at least 3 additional balls in it. The number of those bins is bounded by $\lfloor n/3 \rfloor$ at all time steps $t$, implying a probability of at most 
\begin{align}
\frac{n/3}{n} = \frac{1}{3}
\end{align}
that one such bin is chosen. This implies a probability of at most $3^{-d}$ that all $d$ bins are chosen among those with at least 3 additional bins in them. Thus, the probability that all leaf events are activated is at most $p_l = 3^{-d\cdot l}$, where $l = d^{L}$ is the number of leaves in a witness tree.
